Ubiquitous Digital Technologies and Spatial Structure; an update

### 2020-12-03 paper/2_resubmission/tranos_ioannides_unmarked.Rmd
unmarked

\color{blue}
Kolko [@kolko_death_2000] uses internet diffusion data and identifies a clear complementary link between internet usage and city size.
Interestingly, he identifies higher internet domain densities in remote cities which indicates a substitution effect of the internet for longer-distance non-electronic communications. His results are consistent for different measures of internet diffusion (internet domain density and internet take-up).
Sinai and Waldfogel [@todd_sinai_geography_2004] approach the same question from the consumers’ point of view and find support for complementarity between internet and cities.
They study the link between market size and locally targeted online content and find more online local content in  larger markets.
However, they also obtain evidence for a substitution effect: holding local online content constant, market size has a negative effect on individual connectivity.
Forman et al. [@chris_forman_how_2005] examines whether commercial internet adoption is higher in cities than in rural areas. While the former would indicate a complementarity between internet adoption and cities, the latter would reflect a substitutability relation, according to which the internet is used as a means to offset costs and lack of opportunities related to peripheriality.
Their results indicate that despite internet adoption by firms with more than 100 employees being greater in smaller urban agglomerations, the adoption of more sophisticated internet-based applications is positively related with city size in $2000$.
Sohn, Kim, and Hewings [@jungyul_sohn_information_2003] compare how information technologies are related to urban spatial structure for Chicago and Seoul.
Although they find a clear complementary link for Chicago, this is not the case for Seoul, where information technologies contribute to a more dispersed spatial pattern.
Focusing on the municipalities in the province of Barcelona, Pons-Novell and Viladecans-Marsal [@pons-novell_cities_2006] find a complementary link between individual internet take-up and off-line commercial offerings, but their results cannot safely reject a substitution effect.
Craig, Hoang, and Kohlhase [@craig_does_2016] focus on internet take-up rates across US states during the period $2000-2011$.
Their analysis provides suggestive evidence of a complementary role that internet connectivity performs on urban living.
Ioannides et al. [@ioannides2008effect] use country-level data on city size distributions to examine the impact of fixed line telephony on urban structure $1980-2000$.
Using a panel dataset of spatial dispersion measures, they find robust evidence that an increase in the number of telephone lines per capita encourages the spatial dispersion of population in that they lead to a more concentrated distribution of city sizes.
In addition, by the end of their study period the internet has come into use, but their results with internet usage is more speculative but do show that it goes in the same direction.

\color{blue}
Focusing on rural areas, Partridge et al. [@mark_d_partridge_employment_2008] find no evidence that rural distance penalties in the US have substantially changed since $1970s$ indicating that technological changes including the internet and digital communications have not affected spatial structure. Also interesting are the findings of a more recent study by Kim and Orazem [@kim_broadband_2016] on the economic effects of broadband internet in rural areas. They identify a positive effect on new firm location decisions, with the effect being higher in rural areas with larger population and in rural areas adjacent to a metropolitan area. These results suggest a complementary relation between the internet and agglomeration economies.

\color{blue}
Moving beyond modelling the direct relationship between ICT and spatial structure,
Bekkerman and Gilpin [@bekkerman_high-speed_2013] focused on the role of locally
based information resources using a dataset about the US libraries during the
period $2000-2008$. Their results suggest that internet access increases the
demand and the value of locally accessible information, and such complementarities
are higher in larger metropolitan areas. Anenberg and Kung
[@anenberg_information_2015] also identified a complementary relation between
the internet and consumption variety in cities by focusing on food truck industry
in the US.

@@

\color{blue}
The present paper does not address the issue that adoption of ICT may have
different effects on urban dispersion on a national scale, that is how far are
major urban areas, that is, cities, from one another, versus dispersion within major
urban areas, that is urban sprawl. Such an inquiry would require detailed data
about patterns of how urban sprawl may hamper or promote economic interaction,
given advancements in information technology.
We think that data from the Covid-19 pandemic would likely throw light at this
issue in the future, as it is clear that not all economic activities might
continue in an unencumbered manner; see Dingel and Neiman [@dingel2020many].
\color{black}

@@

The multi-country identification strategy is a two-step approach [@ioannides2008effect].
The first step of our methodology is to estimate the
\color{blue}
Pareto exponent
\color{black}
for a broad sample of countries over time.
\color{blue}
The Pareto exponent
\color{black}
is one of the most widely used measures of spatial dispersion with numerous applications in urban economics and economic geography ([see for example @black_urban_2003; @frenken_theoretical_2007; @giesen_zipfs_2011; @rauch_cities_2013; @ioannides_zipfs_2003; @ioannides_walled_2017; @nitsch_zipf_2005).
This an appropriate measure of dispersion because of the extreme heterogeneity of the city size distribution and the very good fits normally obtained with such estimations^[
\color{blue}
We also replicate our analysis using other measures of dispersion in Appendix S1. Our results remain consistent.].
\color{black}
Skipping details pertaining of the suitability of the approach, which may be found in Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], we proceed with the estimation of the logarithmic form of:

@@

\color{blue}
This leads to an estimate of $\zeta$, known as the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere to for
the remained of the text, even  when $\zeta$ is estimated to be near $1$.
The term Zipf coefficient should be reserved for the special case of
$\zeta=1$.
While much of the literature focuses on the Pareto exponent being
close to $1$, several estimations of city size distributions lead to estimates
of $\zeta$ that are not necessarily equal to $1$, or even near it, in which
case we refer to $\zeta$ as the Pareto, or power law, exponent.
Given that our aim here is to estimate $\zeta$ for a number of countries over
time as a measure of dispersion, equation \ref{eq1} describes the rank of city
$i$ in country $c$ in year $t$:
\color{black}

@@

\color{blue}

@@

\color{black}
\noindent The model enable us to estimate the effects of a vector $X$ of explanatory variables which are included to account for the spatial structure of country $c$ in year $t$.
The main variables of interest here are internet and digital communications variables that include: internet users per $100$ inhabitants, broadband users per $100$ inhabitants, mobile phone users per $100$ inhabitants, and fixed phone users per $100$ inhabitants.
To address a potential omitted variable bias equation \ref{eq3} includes
\color{blue}
country ($\theta_c$) and year ($\delta_t$) fixed effects;
\color{black}
$e_{ct}$ is the error term.
In addition, vector $X$ includes several control variables, whose descriptive statistics together with those for the other variables used to estimate \ref{eq3} are reported in Table \ref{desc.global}.

@@

This section presents the estimation results of \ref{eq3}.
The LHS variable is $\zeta$, the
\color{blue}
Pareto exponent,
\color{black}
as estimated according to the Gabaix and Ibragimov [@gabaix_rank_2011] correction.
The main explanatory variables of interest are provided at the top of Table \ref{ols.global}; namely internet, broadband, mobile and fixed telephony per $100$ habitants, expressed in natural logarithms.
The main variables of interest are introduced successively on their own in the regressions reported in Table \ref{ols.global}.
All regressions include country fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared standard error of the estimated
\color{blue}
Pareto exponent
\color{black}
(see Table \ref{desc.global}) to address potential noise that is carried over from the first part of our identification strategy.
In regards to the interpretation of the estimated coefficients, given that the
\color{blue}
Pareto exponent
\color{black}
has entered the regression not as an absolute value, but instead as a real number a positive coefficient for a RHS variable indicates an impact towards the decrease of the spatial dispersion of population. In other words, a positive coefficient indicates an effect towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies because of the expansion of digital technologies.

@@

Table \ref{ols.global} and \ref{2sls.global} report the estimation results based on the UN urban agglomeration data.
Starting with the former, we note that an agglomerative effect is only detected for the internet users and while a marginally opposing result can be seen for broadband users as indicated by the significant coefficients in columns (1) and (2).
For the telephony variables the estimation of \ref{eq3} did not yield statistically significant coefficients.
Before discussing further these results and the effects of the other control variables, we need to highlight that the main challenge of estimating equation \ref{eq3} is the potential endogenous nature of the share of internet users which might prevent us from being able to infer a truly causal effect.
Endogeneity might be an issue here as spatial structure, which is proxied by the
\color{blue}
Pareto exponent,
\color{black}
might be affected by another source, which also affects internet penetration.
For instance, economic development might affect the concentration of population in large cities and at the same time enable more people to go online.
If we do not address this issue, the coefficient for the main variable of interest will capture potential effects that internet penetration has on spatial structure, but also potential reverse causality effects that spatial structure might generate on internet penetration.
To overcome this potential problem, Table \ref{2sls.global} reports estimates of equation \ref{eq3} using two-stage least squares (2SLS) with instrumental variables (IVs).
The latter are variables which are correlated with our endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any -- of the internet and digital communications on spatial structure.
At a first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs and the other control variables are used instead of the endogenous variable to estimate eq. \ref{eq3}.
A significant effect will verify the causal impact of the internet and digital communication usage on spatial structure.

@@

The main challenge for such an exercise is to find a valid IV.
In our case, the challenge is even bigger as we need to find a variable which is not only fit for purpose, but it also contain as few missing values as possible in order to retain the same number of observations as the OLS models.
This is particularly difficult for a diverse sample of countries as the one we are dealing with here.
We propose here the percentage of female participation in the labor force, which is directly related to digital infrastructure, but not to spatial structure (see Table \ref{desc.global} for the descriptive statistics).
\color{blue}
The literature provides empirical and theoretical evidence that ICT are correlated with female participation in labour markets mostly because they enable teleworking [@dettling2017; @arntz2019working; @watson2018digitalisation; @Lechman2020; @Valberg2020]
\color{black}
At the same time, female labor participation is not related with spatial structure especially given the diverse sample of countries.
<!-- Firstly, the correlation coefficient between the IV and the Zipf coefficient is very low ($-0.14$). -->
Although one might think that increased female labor participation may lead to relocation of households to large cities and therefore affect the spatial structure of cities, such a process varies a lot within our dataset.
Even if this might be true for developed countries with mature urban systems for which a location within a large urban center might provide opportunities for both male and female workers to find jobs, our dataset also includes countries, $41$ per cent of the GDP of which is attributed to primary activities (see Table \ref{desc.global}).
So, female labor participation might also be related to economic activities located outside large urban centers.
In terms of the relevant tests presented in Table \ref{2sls.global} the weak identification first stage F-test exceeds by far all the Stock-Yogo critical values for all, but the mobile telephony regression.
Hence, we are confident to interpret the effects of internet, broadband internet and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that increases in internet and broadband internet usage have resulted to a decrease of the spatial dispersion of population for the time period and for the panel of countries included in our data.
In other words, increases in internet and digital communications resulted to national urban systems which are less uniform in terms of city sizes and are characterized by higher population concentrations in larger cities.
The exact opposite appears to happen for fixed telephony, which our results indicate that has led to an increase of spatial dispersion of population.
\color{blue}
We expect that the IV estimates capture the local average treatment effect
[@angrist1995identification] of these labour markets, which are characterised by
more active engagement with the internet, and, therefore, experience a larger effect
on the Pareto exponents.
\color{black}
Importantly, our identification strategy enables us to address potential reverse causality issues and treat the results of Table \ref{2sls.global} as causal. Hence, the main finding of our multi-country, global analysis based on data for urban agglomerations of at least 300,000 inhabitants is that internet and broadband internet usage appear to act in favor of agglomeration economies and result to urban systems with more dominant cities on the top of the urban hierarchy.
On the contrary, fixed telephony usage acted in favour of centrifugal forces.
Our identification strategy does not enable us to make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and telephony penetration have further enhanced agglomeration forces, at least for large urban agglomerations, while the opposite happens for fixed telephony.
These results are robust against endogeneity issues, but are limited to urban agglomeration included in our data.
Indeed, for the estimation of the
\color{blue}
Pareto exponent
\color{black}
we only included urban agglomeration of $300,000$ inhabitants or more due to data availability.
Hence, the above estimations cannot verify whether such an effect is also valid for smaller cities.
Interestingly, the above results regarding the role of fixed telephony are in agreement with previous results from [@ioannides2008effect] [@tranos2020ict] which used a smaller panel of countries but without constraints regarding the city size.
However, our results do not support the findings of [@tranos2020ict] regarding internet usage.

\color{blue}
It is indeed puzzling that the results on Table \ref{2sls.global} that the
effects on urban concentration for internet users and broadband users, on the
one hand, and mobile and fixed telephony users are in opposite directions.
We wish to broach this subject in the context of some related literature,
namely [@calabrese2011interplay] and [@buchel2020cities].
The latter supports the notion that ICT favors agglomeration in the sense that
in spite of their convenience modern technologies do no make up for the need of
proximity in  social interactions.
The former supports the notion people who communicate more frequently are likely
to be near one another.
ICT, as viewed through such granular studies of individuals' communications
do not make up for physical distance; individuals nearer one another
communicate more.  
\color{black}

@@

The LHS variable we use here is the outcome of equation \ref{eq5}.
\color{blue}
We use this variable to estimate \ref{eq4} with OLS.
Additionally, we employ a normalized form of the dependent variable, so as that
variable is bounded between $0$ and $1$ and with a mean of  $0.5$^[$$Normalized\:difference\:in\:ranks_{i} = \frac{difference\:in\:ranks_{i} + N_{cities}}{2xN_{cities}}.$$].
This transformation enables us to estimate \ref{eq4}
with a quasibinomial GLM estimator given that the original form of our LHS
variable does not vary continuously, but is instead defined as a difference between
two count variables, which may also assume negative values.
\color{blue}
The results of the different estimations remain qualitatively the same regardless of the form of the LHS variable we use.
Table \ref{ols.usa} reports the results of the estimation of equation \ref{eq4} using OLS.
The sign and significance of the main variable of interest -- percentage of population with computer and broadband connection -- verifies our global level results.
That is, an increase in internet usage improves the position of a city in the US urban system.
The results are consistent among the different estimators that is OLS and GLM.
In addition, Table \ref{ols.usa} also reports the estimates of interaction effects between the share of population with broadband connection with population and population density.
Although the interaction terms are not significant, the main effect remains qualitatively unchanged.

@@

Although not directly comparable, the above estimations are in accordance with our global model. Moreover, the potential presence of endogeneity might be a problem here as well.
Therefore, Table \ref{2sls.usa} reports 2SLS estimations.
In addition to the strategy employed for the cross-country analysis -- that is the inclusion of one instrument for which we have strong reasons to believe is uncorrelated with the error term -- data availability allows us to employ a second IV in order to estimate the Sargan over-identification restrictions test (Column 2).
The main instrument we propose here is Bachelor’s degrees per inhabitant in $2005$^[
\color{blue}
Due to differences in scale, time and the endogenous variables, different IVs
have been unutilized for the different sections of the paper.
\color{black}].
Even if the quality of human capital affect the population growth of a city $10$ years later, our LHS variable adopts a systemic understanding of the US urban system as it measures the relative position of a city within the overall urban system instead of its population growth.
In other words, a city might experience population growth between two periods, but if other cities have also experienced population growth, this might not affect its relative position within the urban system.
Hence, we do not expect that this IV affects the LHS variable.
<!-- In addition, the correlation coefficient between these two variables is still very low ($r=0.24$). -->
Moreover, when we add the second IV, which in this case is the commute time in minutes also in $2005$, we fail to reject the null hypothesis of the Sargan test, something which also adds the validity of our strategy.
Furthermore, our results do not suffer from weak identification according to the relevant tests in Table \ref{2sls.usa}.

@@

In total, the estimates of Table \ref{2sls.usa} enable us to identify a causal effect of the share of households with internet connection on the position of a city in the urban system.
Specifically, an increase in internet penetration led to an improvement of the position of a city in the US urban hierarchy for the cities and the time period included in our analysis.
\color{blue}
Everything else being equal, if a city had experienced an increase of 10 per cent
of internet penetration, this increase would have improved its relative position by $9$ places.
\color{black}
As in the global model reported in [Section 4](#sec4), internet penetration appears to work in favor of agglomeration externalities.
In regards to the control variables, we can identify a negative effect of income, which is consistent with the effect of GDP per capita in the global model presented in Table \ref{2sls.global}.
During the study period, unemployment rates and the share of white population negatively affected the ranking of micropolitan and metropolitan areas in the US as reflected in the relevant regressions.

>>>>

<!-- <!-- \color{blue} --> -->
Kolko [@kolko_death_2000] uses internet diffusion data and identifies a clear complementary link between internet usage and city size.
Interestingly, he identifies higher internet domain densities in remote cities which indicates a substitution effect of the internet for longer-distance non-electronic communications. His results are consistent for different measures of internet diffusion (internet domain density and internet take-up).
Sinai and Waldfogel [@todd_sinai_geography_2004] approach the same question from the consumers’ point of view and find support for complementarity between internet and cities.
They study the link between market size and locally targeted online content and find more online local content in  larger markets.
However, they also obtain evidence for a substitution effect: holding local online content constant, market size has a negative effect on individual connectivity.
Forman et al. [@chris_forman_how_2005] examines whether commercial internet adoption is higher in cities than in rural areas. While the former would indicate a complementarity between internet adoption and cities, the latter would reflect a substitutability relation, according to which the internet is used as a means to offset costs and lack of opportunities related to peripheriality.
Their results indicate that despite internet adoption by firms with more than 100 employees being greater in smaller urban agglomerations, the adoption of more sophisticated internet-based applications is positively related with city size in $2000$.
Sohn, Kim, and Hewings [@jungyul_sohn_information_2003] compare how information technologies are related to urban spatial structure for Chicago and Seoul.
Although they find a clear complementary link for Chicago, this is not the case for Seoul, where information technologies contribute to a more dispersed spatial pattern.
Focusing on the municipalities in the province of Barcelona, Pons-Novell and Viladecans-Marsal [@pons-novell_cities_2006] find a complementary link between individual internet take-up and off-line commercial offerings, but their results cannot safely reject a substitution effect.
Craig, Hoang, and Kohlhase [@craig_does_2016] focus on internet take-up rates across US states during the period $2000-2011$.
Their analysis provides suggestive evidence of a complementary role that internet connectivity performs on urban living.
Ioannides et al. [@ioannides2008effect] use country-level data on city size distributions to examine the impact of fixed line telephony on urban structure $1980-2000$.
Using a panel dataset of spatial dispersion measures, they find robust evidence that an increase in the number of telephone lines per capita encourages the spatial dispersion of population in that they lead to a more concentrated distribution of city sizes.
In addition, by the end of their study period the internet has come into use, but their results with internet usage is more speculative but do show that it goes in the same direction.

<!-- <!-- \color{blue} --> -->
Focusing on rural areas, Partridge et al. [@mark_d_partridge_employment_2008] find no evidence that rural distance penalties in the US have substantially changed since $1970s$ indicating that technological changes including the internet and digital communications have not affected spatial structure. Also interesting are the findings of a more recent study by Kim and Orazem [@kim_broadband_2016] on the economic effects of broadband internet in rural areas. They identify a positive effect on new firm location decisions, with the effect being higher in rural areas with larger population and in rural areas adjacent to a metropolitan area. These results suggest a complementary relation between the internet and agglomeration economies.

<!-- <!-- \color{blue} --> -->
Moving beyond modelling the direct relationship between ICT and spatial structure,
Bekkerman and Gilpin [@bekkerman_high-speed_2013] focused on the role of locally
based information resources using a dataset about the US libraries during the
period $2000-2008$. Their results suggest that internet access increases the
demand and the value of locally accessible information, and such complementarities
are higher in larger metropolitan areas. Anenberg and Kung
[@anenberg_information_2015] also identified a complementary relation between
the internet and consumption variety in cities by focusing on food truck industry
in the US.

@@

<!-- \color{blue} -->
The present paper does not address the issue that adoption of ICT may have
different effects on urban dispersion on a national scale, that is how far are
major urban areas, that is, cities, from one another, versus dispersion within major
urban areas, that is urban sprawl. Such an inquiry would require detailed data
about patterns of how urban sprawl may hamper or promote economic interaction,
given advancements in information technology.
We think that data from the Covid-19 pandemic would likely throw light at this
issue in the future, as it is clear that not all economic activities might
continue in an unencumbered manner; see Dingel and Neiman [@dingel2020many].
\color{black}

@@

The multi-country identification strategy is a two-step approach [@ioannides2008effect].
The first step of our methodology is to estimate the
<!-- \color{blue} -->
Pareto exponent
\color{black}
for a broad sample of countries over time.
<!-- \color{blue} -->
The Pareto exponent
\color{black}
is one of the most widely used measures of spatial dispersion with numerous applications in urban economics and economic geography ([see for example @black_urban_2003; @frenken_theoretical_2007; @giesen_zipfs_2011; @rauch_cities_2013; @ioannides_zipfs_2003; @ioannides_walled_2017; @nitsch_zipf_2005).
This an appropriate measure of dispersion because of the extreme heterogeneity of the city size distribution and the very good fits normally obtained with such estimations^[
<!-- \color{blue} -->
We also replicate our analysis using other measures of dispersion in Appendix S1. Our results remain consistent.].
\color{black}
Skipping details pertaining of the suitability of the approach, which may be found in Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], we proceed with the estimation of the logarithmic form of:

@@

<!-- \color{blue} -->
This leads to an estimate of $\zeta$, known as the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere to for
the remained of the text, even  when $\zeta$ is estimated to be near $1$.
The term Zipf coefficient should be reserved for the special case of
$\zeta=1$.
While much of the literature focuses on the Pareto exponent being
close to $1$, several estimations of city size distributions lead to estimates
of $\zeta$ that are not necessarily equal to $1$, or even near it, in which
case we refer to $\zeta$ as the Pareto, or power law, exponent.
Given that our aim here is to estimate $\zeta$ for a number of countries over
time as a measure of dispersion, equation \ref{eq1} describes the rank of city
$i$ in country $c$ in year $t$:
\color{black}

@@

<!-- \color{blue} -->

@@

\color{black}
\noindent The model enable us to estimate the effects of a vector $X$ of explanatory variables which are included to account for the spatial structure of country $c$ in year $t$.
The main variables of interest here are internet and digital communications variables that include: internet users per $100$ inhabitants, broadband users per $100$ inhabitants, mobile phone users per $100$ inhabitants, and fixed phone users per $100$ inhabitants.
To address a potential omitted variable bias equation \ref{eq3} includes
<!-- \color{blue} -->
country ($\theta_c$) and year ($\delta_t$) fixed effects;
\color{black}
$e_{ct}$ is the error term.
In addition, vector $X$ includes several control variables, whose descriptive statistics together with those for the other variables used to estimate \ref{eq3} are reported in Table \ref{desc.global}.

@@

This section presents the estimation results of \ref{eq3}.
The LHS variable is $\zeta$, the
<!-- \color{blue} -->
Pareto exponent,
\color{black}
as estimated according to the Gabaix and Ibragimov [@gabaix_rank_2011] correction.
The main explanatory variables of interest are provided at the top of Table \ref{ols.global}; namely internet, broadband, mobile and fixed telephony per $100$ habitants, expressed in natural logarithms.
The main variables of interest are introduced successively on their own in the regressions reported in Table \ref{ols.global}.
All regressions include country fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared standard error of the estimated
<!-- \color{blue} -->
Pareto exponent
\color{black}
(see Table \ref{desc.global}) to address potential noise that is carried over from the first part of our identification strategy.
In regards to the interpretation of the estimated coefficients, given that the
<!-- \color{blue} -->
Pareto exponent
\color{black}
has entered the regression not as an absolute value, but instead as a real number a positive coefficient for a RHS variable indicates an impact towards the decrease of the spatial dispersion of population. In other words, a positive coefficient indicates an effect towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies because of the expansion of digital technologies.

@@

Table \ref{ols.global} and \ref{2sls.global} report the estimation results based on the UN urban agglomeration data.
Starting with the former, we note that an agglomerative effect is only detected for the internet users and while a marginally opposing result can be seen for broadband users as indicated by the significant coefficients in columns (1) and (2).
For the telephony variables the estimation of \ref{eq3} did not yield statistically significant coefficients.
Before discussing further these results and the effects of the other control variables, we need to highlight that the main challenge of estimating equation \ref{eq3} is the potential endogenous nature of the share of internet users which might prevent us from being able to infer a truly causal effect.
Endogeneity might be an issue here as spatial structure, which is proxied by the
<!-- \color{blue} -->
Pareto exponent,
\color{black}
might be affected by another source, which also affects internet penetration.
For instance, economic development might affect the concentration of population in large cities and at the same time enable more people to go online.
If we do not address this issue, the coefficient for the main variable of interest will capture potential effects that internet penetration has on spatial structure, but also potential reverse causality effects that spatial structure might generate on internet penetration.
To overcome this potential problem, Table \ref{2sls.global} reports estimates of equation \ref{eq3} using two-stage least squares (2SLS) with instrumental variables (IVs).
The latter are variables which are correlated with our endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any -- of the internet and digital communications on spatial structure.
At a first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs and the other control variables are used instead of the endogenous variable to estimate eq. \ref{eq3}.
A significant effect will verify the causal impact of the internet and digital communication usage on spatial structure.

@@

The main challenge for such an exercise is to find a valid IV.
In our case, the challenge is even bigger as we need to find a variable which is not only fit for purpose, but it also contain as few missing values as possible in order to retain the same number of observations as the OLS models.
This is particularly difficult for a diverse sample of countries as the one we are dealing with here.
We propose here the percentage of female participation in the labor force, which is directly related to digital infrastructure, but not to spatial structure (see Table \ref{desc.global} for the descriptive statistics).
<!-- \color{blue} -->
The literature provides empirical and theoretical evidence that ICT are correlated with female participation in labour markets mostly because they enable teleworking [@dettling2017; @arntz2019working; @watson2018digitalisation; @Lechman2020; @Valberg2020]
\color{black}
At the same time, female labor participation is not related with spatial structure especially given the diverse sample of countries.
<!-- Firstly, the correlation coefficient between the IV and the Zipf coefficient is very low ($-0.14$). -->
Although one might think that increased female labor participation may lead to relocation of households to large cities and therefore affect the spatial structure of cities, such a process varies a lot within our dataset.
Even if this might be true for developed countries with mature urban systems for which a location within a large urban center might provide opportunities for both male and female workers to find jobs, our dataset also includes countries, $41$ per cent of the GDP of which is attributed to primary activities (see Table \ref{desc.global}).
So, female labor participation might also be related to economic activities located outside large urban centers.
In terms of the relevant tests presented in Table \ref{2sls.global} the weak identification first stage F-test exceeds by far all the Stock-Yogo critical values for all, but the mobile telephony regression.
Hence, we are confident to interpret the effects of internet, broadband internet and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that increases in internet and broadband internet usage have resulted to a decrease of the spatial dispersion of population for the time period and for the panel of countries included in our data.
In other words, increases in internet and digital communications resulted to national urban systems which are less uniform in terms of city sizes and are characterized by higher population concentrations in larger cities.
The exact opposite appears to happen for fixed telephony, which our results indicate that has led to an increase of spatial dispersion of population.
<!-- \color{blue} -->
We expect that the IV estimates capture the local average treatment effect
[@angrist1995identification] of these labour markets, which are characterised by
more active engagement with the internet, and, therefore, experience a larger effect
on the Pareto exponents.
\color{black}
Importantly, our identification strategy enables us to address potential reverse causality issues and treat the results of Table \ref{2sls.global} as causal. Hence, the main finding of our multi-country, global analysis based on data for urban agglomerations of at least 300,000 inhabitants is that internet and broadband internet usage appear to act in favor of agglomeration economies and result to urban systems with more dominant cities on the top of the urban hierarchy.
On the contrary, fixed telephony usage acted in favour of centrifugal forces.
Our identification strategy does not enable us to make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and telephony penetration have further enhanced agglomeration forces, at least for large urban agglomerations, while the opposite happens for fixed telephony.
These results are robust against endogeneity issues, but are limited to urban agglomeration included in our data.
Indeed, for the estimation of the
<!-- \color{blue} -->
Pareto exponent
\color{black}
we only included urban agglomeration of $300,000$ inhabitants or more due to data availability.
Hence, the above estimations cannot verify whether such an effect is also valid for smaller cities.
Interestingly, the above results regarding the role of fixed telephony are in agreement with previous results from [@ioannides2008effect] [@tranos2020ict] which used a smaller panel of countries but without constraints regarding the city size.
However, our results do not support the findings of [@tranos2020ict] regarding internet usage.

<!-- \color{blue} -->
It is indeed puzzling that the results on Table \ref{2sls.global} that the
effects on urban concentration for internet users and broadband users, on the
one hand, and mobile and fixed telephony users are in opposite directions.
We wish to broach this subject in the context of some related literature,
namely [@calabrese2011interplay] and [@buchel2020cities].
The latter supports the notion that ICT favors agglomeration in the sense that
in spite of their convenience modern technologies do no make up for the need of
proximity in  social interactions.
The former supports the notion people who communicate more frequently are likely
to be near one another.
ICT, as viewed through such granular studies of individuals' communications
do not make up for physical distance; individuals nearer one another
communicate more.  
\color{black}

@@

The LHS variable we use here is the outcome of equation \ref{eq5}.
<!-- \color{blue} -->
We use this variable to estimate \ref{eq4} with OLS.
Additionally, we employ a normalized form of the dependent variable, so as that
variable is bounded between $0$ and $1$ and with a mean of  $0.5$^[$$Normalized\:difference\:in\:ranks_{i} = \frac{difference\:in\:ranks_{i} + N_{cities}}{2xN_{cities}}.$$].
This transformation enables us to estimate \ref{eq4}
with a quasibinomial GLM estimator given that the original form of our LHS
variable does not vary continuously, but is instead defined as a difference between
two count variables, which may also assume negative values.
<!-- \color{blue} -->
The results of the different estimations remain qualitatively the same regardless of the form of the LHS variable we use.
Table \ref{ols.usa} reports the results of the estimation of equation \ref{eq4} using OLS.
The sign and significance of the main variable of interest -- percentage of population with computer and broadband connection -- verifies our global level results.
That is, an increase in internet usage improves the position of a city in the US urban system.
The results are consistent among the different estimators that is OLS and GLM.
In addition, Table \ref{ols.usa} also reports the estimates of interaction effects between the share of population with broadband connection with population and population density.
Although the interaction terms are not significant, the main effect remains qualitatively unchanged.

@@

Although not directly comparable, the above estimations are in accordance with our global model. Moreover, the potential presence of endogeneity might be a problem here as well.
Therefore, Table \ref{2sls.usa} reports 2SLS estimations.
In addition to the strategy employed for the cross-country analysis -- that is the inclusion of one instrument for which we have strong reasons to believe is uncorrelated with the error term -- data availability allows us to employ a second IV in order to estimate the Sargan over-identification restrictions test (Column 2).
The main instrument we propose here is Bachelor’s degrees per inhabitant in $2005$^[
<!-- \color{blue} -->
Due to differences in scale, time and the endogenous variables, different IVs
have been unutilized for the different sections of the paper.
\color{black}].
Even if the quality of human capital affect the population growth of a city $10$ years later, our LHS variable adopts a systemic understanding of the US urban system as it measures the relative position of a city within the overall urban system instead of its population growth.
In other words, a city might experience population growth between two periods, but if other cities have also experienced population growth, this might not affect its relative position within the urban system.
Hence, we do not expect that this IV affects the LHS variable.
<!-- In addition, the correlation coefficient between these two variables is still very low ($r=0.24$). -->
Moreover, when we add the second IV, which in this case is the commute time in minutes also in $2005$, we fail to reject the null hypothesis of the Sargan test, something which also adds the validity of our strategy.
Furthermore, our results do not suffer from weak identification according to the relevant tests in Table \ref{2sls.usa}.

@@

In total, the estimates of Table \ref{2sls.usa} enable us to identify a causal effect of the share of households with internet connection on the position of a city in the urban system.
Specifically, an increase in internet penetration led to an improvement of the position of a city in the US urban hierarchy for the cities and the time period included in our analysis.
<!-- \color{blue} -->
Everything else being equal, if a city had experienced an increase of 10 per cent
of internet penetration, this increase would have improved its relative position by $9$ places.
\color{black}
As in the global model reported in [Section 4](#sec4), internet penetration appears to work in favor of agglomeration externalities.
In regards to the control variables, we can identify a negative effect of income, which is consistent with the effect of GDP per capita in the global model presented in Table \ref{2sls.global}.
During the study period, unemployment rates and the share of white population negatively affected the ranking of micropolitan and metropolitan areas in the US as reflected in the relevant regressions.
